\textsc{geepers}: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-12-19 writeUps/paper.tex
submit v1

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}
\usepackage{multibib}

@@

\newcites{supp}{Supplementary References}

@@

\author[Sales and Vanacore]{Adam C Sales}

\author{Kirk Vanacore}
\address{Worcester Polytechnic Institute,
Worcester, MA,
USA.}
\email{asales@wpi.edu}

@@

In a separate, if related, line of research, \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on a general estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
It turns out that, under certain general conditions, principal effects can be estimated with a simple, two-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores and estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``\textsc{geepers}'': generalized estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability, and admits to sandwich-style standard error estimates, which we provide in the appendix.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are non-parametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
% When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
% In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$% that do not assume the exclusion restriction
---normal mixture modeling and weighting---with some comments in-between on the role for covariates.

@@

\subsection{Simulation Design}
The simulation study was conducted in \texttt{R} \citep{rcite} and Stan \citep{rstan}, and full replication code is available at \url{www.github.com/[Redacted]}.
\subsubsection{Data Generation}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth,clip]{../simFigs/boxplots.pdf}
  \caption{Violin and scatterplots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}

Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatterplots, showing estimation error for \textsc{geepers}, the mixture estimator, and the \textsc{psw} estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates under a somewhat wider set of circumstances when $\muc1=0.3$. More complete results, including root-mean-squared-error estimates, are in the online appendix. 
The leftmost panel of Figure \ref{fig:boxplots} and the top line of Table \ref{tab:coverage} show results under conditions favorable to both the mixture model and to \textsc{geepers}---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
\textsc{geepers} was roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it was much noisier, and when $\alpha=0.5$ it was only slightly less precise.
The mixture estimator's bias depended on $\muc1$---when $\muc1=0$, the bias was slight, and when $\muc1=0.3$ the bias was more substantial. This is somewhat surprising, because when $\muc1=0$, there was no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
The \textsc{psw} estimator was also biased---due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:PI}---inherent in the data generating model. On the other hand, the \textsc{psw} estimator had easily the lowest sampling variance of the three estimators.
The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:coverage} show a similar set of results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
In this case, \textsc{geepers} and \textsc{psw} estimator, which do not assume normality, behaved in a similar fashion as in the normal case, but the mixture estimator showed a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.
Credible intervals from the mixture model exhibited severe undercoverage, while \textsc{geepers} confidence intervals achieved at least their nominal level.

@@

In this experiment, the ability to bottom out was only one of several differences between the treatment conditions. One way to disentangle the role of bottoming out in any treatment effect would be to estimate average treatment effects of being assigned to ASSISTments, relative to any of the other three, separately for subjects who would (if assigned to ASSISTments) bottom out frequently, and for subjects who would not. 
If the availability of bottom-out hints plays an important positive role in ASSISTments's efficacy, we might expect larger effects of condition on students who would bottom out frequently than for those who would not; if bottom-out hints are harmful, we might expect the opposite. 
%If hints (that do not supply the correct answer) and error messages are helpful, but the ability to bottom out is harmful to learning, then we might expect to estimate a positive effect of being assigned to the feedback condition for students who do not bottom out, but a negative effect for students who do. If hints are helpful for learning only if they can be read as worked examples, then we might expect no effect of condition on students who generally do not bottom out, but a positive effect for students who do. 
(These hypothetical results may be masked by other treatment effect heterogeneity between students who would or wouldn't bottom out, but the results could still be informative.)

@@

We also had access to data on a number of baseline covariates, including prior achievement, demographics, baseline measurements of student attitudes towards math and an indicator for whether students began the school year in remote or in-person instruction.
Students with missing pretest scores were dropped from the analysis; missing data in other covariates was imputed with a Random Forest algorithm \citep{missForest}.

@@

Due to privacy regulations, the data are not publicly available; however, interested analysts may access anonymized data by emailing the study's principal investigator at [REDACTED] and signing a privacy agreement.
Replication code for the analysis is available at \url{www.github.com/[Redacted]}.

@@

We estimated principal scores using logistic regression of observed $S$ on school fixed effects and a set of baseline variables chosen with a combination of expert judgement and backwards selection based on the AIC, and evaluated model fit with binned residual plots \citep{gelman2006data}.
The model was fit to data from students in the ASSISTments group, for whom bottom-out status ($S$) was observed; hence, the same model fit was used for all three treatment contrasts.
The model was fairly successful in distinguishing bottom-outers from non-bottom-outers---the AUC, evaluated using out-of-sample predictions in a 10-fold cross-validation, was roughly 0.795. That is, in roughly 80\% of bottom-outer/non-bottom-outer pairs of subjects, the bottom-outer will have a higher predicted probability from the model.

@@

Figure \ref{fig:compare} shows principal effect estimates from \textsc{geepers} compared to analogous estimates using Bayesian mixture modeling and \textsc{psw}, alongside estimates of the overall average treatment effects, labeled ``ATE.''.
All three sets of estimates relied on the same principal score model described above, in Section \ref{sec:psMod}, though in the mixture model it was fit simultaneously with effect estimation using a Markov Chain Monte Carlo algorithm \citep{rstan}.
Outcome regressions in \textsc{geepers} and the mixture model included all of the covariates that were in the principal score model, in addition to students' pretest scores; also, the models substituted class fixed effects for school fixed effects.

The Bayesian mixture model assumed that outcomes were normally distributed, conditional on covariates (the same as in Section \ref{sec:outMod}) and principal stratum, with a residual standard deviation that varied between principal strata. This modeling assumption is necessarily false, since post-test scores were integers between 0 and 10. All model parameters with support in $\mathbb{R}$ were given standard normal priors, while standard deviation parameters were given half-standard normal priors.

@@

\section{Discussion}\label{sec:discussion}
\textsc{geepers} is a straightforward approach to principal effect estimation under strong monotonicity, built on widely-used regression models, which is more robust---though less precise---than alternative approaches under a wide array of scenarios.

There is good reason to hope that extensions to \textsc{geepers} may be straightforward.
First, \textsc{geepers} may be extended to situations in which there is two-way noncompliance, and only a weaker form of monotonicity (Assumption \ref{ass:sm}) holds.
For instance, in truncation by death problems \citep[e.g.][]{zhangRubin,ding2011}, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, and interest is is in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
The scenario of truncation by death with (weak) monotonicity is broadly similar to the strong monotonicity, one-way noncompliance situation that this paper discussed.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

\bibliographystyle{plainnat}
\bibliography{MOM}
\appendix

@@

Replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations $\sum_{i=1}^{n_C}\Psi_i=\bm{0}$ or 
\begin{equation}\label{eq:estEq1}
\sum_{i=1}^{n_C}\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}=\begin{pmatrix} 0\\0\\0\\0\end{pmatrix}
\end{equation}
These are equivalent to the estimating equations for OLS model \eqref{eq:regression0} with $\beta_0=\muc0$, $\beta_1=\muc1-\muc0$, $\beta_2=\mut0-\muc0$, and $\beta_3=\mut1-\mut0-\muc1-\muc0$.
Therefore, under standard OLS regularity conditions the estimated parameter vector $\bm{\hat{\beta}}$ is consistent, completing the proof.

@@

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEst}
  \begin{split}
    \heff0_{int}&\equiv \hat{\beta}_2+\bm{\hat{\gamma}_3}'\overline{\bxy}_{Z=1,S=0}\\
    \heff1_{int}&\equiv \hat{\beta}_2+\hat{\beta}_3+(\bm{\hat{\gamma}_3}+\bm{\hat{\gamma}_4})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\heff0_{G}$ and $\heff1_{int}$ are M-estimators, with $\heff0_{int}\rightarrow_p\eff0$ and $\heff1_{G}\rightarrow_p\eff1$ as $n\rightarrow\infty$.
  Under suitible regularity conditions, they are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

@@

\begin{proof}
First of all, by \eqref{eq:interaction}, 
\begin{equation*}
  \EE[Y_T-Y_C|\st=0]=\EE[Y|Z=1,\st=0]-\EE[Y|Z=1,\st=0]=\beta_2+\bm{\gamma_3}'\EE[\bxy|\st=0]
\end{equation*}
and
\begin{equation*}
  \EE[Y_T-Y_C|\st=1]=\EE[Y|Z=1,\st=1]-\EE[Y|Z=1,\st=1]=\beta_2+\beta_3+(\bm{\gamma_3}'+\bm{\gamma_4}')\EE[\bxy|\st=1]
\end{equation*}
Furthermore, $\overline{\bxy}_{Z=1,S=0}\rightarrow \EE[\bxy|\st=0]$ and $\overline{\bxy}_{Z=1,S=1}\rightarrow \EE[\bxy|\st=1]$ as $n\rightarrow \infty$.

@@

First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ (due to randomization)}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}
implying that, according to \eqref{eq:interaction},
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\beta_1\pp+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}\pp\\
                  &=\beta_0+\beta_1R+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}R\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_2+(\beta_1+\beta_3)S+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy S\\
    &=\beta_0+\beta_2+(\beta_1+\beta_3)R+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy R
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]&=\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  &=\beta_0+\beta_1\EE[R]+\beta_2\EE[Z]+\beta_3\EE[ZR]+\bm{\gamma_1}'\EE[\bxy]+\bm{\gamma_2}'\EE[\bxy R]+\bm{\gamma_3}'\EE[\bxy Z]+\bm{\gamma_4}\EE[\bxy RZ]
\end{align*}

@@

\clearpage
\section{Simulation Results}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{simFigs/alphaAUC.pdf}
    \caption{Estimated AUC values from principal score models for each value of $\alpha$ in the simulation, with $n=500$.}
    \label{fig:alphaAUC}
\end{figure}

>>>>

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}
%\usepackage{multibib}

@@

%\newcites{supp}{Supplementary References}

@@

\author[Sales et al.]{Adam C Sales}

\author{Kirk P Vanacore}
\author{Erin R Ottmar}

\address{Worcester Polytechnic Institute,
Worcester, MA,
USA.}
\email{asales@wpi.edu}

@@

On the other hand, \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on a general estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
It turns out that, under certain general conditions, principal effects can be estimated with a simple, two-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores and estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``\textsc{geepers}'': generalized estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability, and admits to sandwich-style standard error estimates, which we provide in the appendix.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are nonparametrically  identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
% When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
% In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$% that do not assume the exclusion restriction
---normal mixture modeling and weighting---with some comments in-between on the role for covariates.

@@

\subsection{Simulation Design}
The simulation study was conducted in \texttt{R} \citep{rcite} and Stan \citep{rstan}, and full replication code is available at \url{https://github.com/adamSales/psGee}.
\subsubsection{Data Generation}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth,clip]{../simFigs/boxplots.pdf}
  \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}

Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation error for \textsc{geepers}, the mixture estimator, and the \textsc{psw} estimator of $\eff1$ under various sets of conditions, and Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates under a somewhat wider set of circumstances when $\muc1=0.3$. More complete results, including root-mean-squared-error estimates, are in the online appendix. 
The leftmost panel of Figure \ref{fig:boxplots} and the top line of Table \ref{tab:coverage} show results under conditions favorable to both the mixture model and to \textsc{geepers}---normal residuals and no interactions in the data generating model between covariates and either $S$ or $Z$.
\textsc{geepers} was roughly unbiased, but less precise than the other two methods---when $\alpha=0.2$ it was much noisier, and when $\alpha=0.5$ it was only slightly less precise.
The mixture estimator's bias depended on $\muc1$---when $\muc1=0$, the bias was slight, and when $\muc1=0.3$ the bias was more substantial. This is somewhat surprising, because when $\muc1=0$, there was no separation between the principal strata in the control group, and previous literature \citep{griffin2008application} has identified this scenario as a particular challenge for mixture modeling.
The \textsc{psw} estimator was also biased---due, presumably, to the violation of Principal Ignorability, Assumption \ref{ass:PI}---inherent in the data generating model. On the other hand, the \textsc{psw} estimator had easily the lowest sampling variance of the three estimators.
The middle panel of Figure \ref{fig:boxplots} and the fifth line of Table \ref{tab:coverage} show a similar set of results for when the residuals in \eqref{eq:y-sim} are uniformly distributed.
In this case, \textsc{geepers} and \textsc{psw} estimator, which do not assume normality, behaved in a similar fashion as in the normal case, but the mixture estimator showed a bimodal pattern--sometimes over-estimating $\eff1$ and sometimes underestimating, but rarely estimating $\eff1$ accurately.
Credible intervals from the mixture model exhibited severe undercoverage, while \textsc{geepers} confidence intervals achieved at least their nominal level.

@@

In this experiment, the ability to bottom out was only one of several differences between the treatment conditions. One way to disentangle the role of bottoming out in any treatment effect would be to estimate average treatment effects of being assigned to ASSISTments, relative to any of the other three, separately for subjects who would (if assigned to ASSISTments) bottom out frequently, and for subjects who would not. 
If the availability of bottom-out hints plays an important positive role in ASSISTments' efficacy, we might expect larger effects of condition on students who would bottom out frequently than for those who would not; if bottom-out hints are harmful, we might expect the opposite. 
%If hints (that do not supply the correct answer) and error messages are helpful, but the ability to bottom out is harmful to learning, then we might expect to estimate a positive effect of being assigned to the feedback condition for students who do not bottom out, but a negative effect for students who do. If hints are helpful for learning only if they can be read as worked examples, then we might expect no effect of condition on students who generally do not bottom out, but a positive effect for students who do. 
(These hypothetical results may be masked by other treatment effect heterogeneity between students who would or wouldn't bottom out, but the results could still be informative.)

@@

We also had access to data on a number of baseline covariates, including prior achievement, demographics, baseline measurements of student attitudes towards math and an indicator for whether students began the school year in remote or in-person instruction.
Students with missing pretest scores were dropped from the analysis; missing data in other covariates was imputed with a Random Forest algorithm \citep{missForest}. Summary statistics are displayed in the online appendix.

@@

Due to privacy regulations, the data are not publicly available; however, interested analysts may access anonymized data by emailing the study's principal investigator at erottmar@wpi.edu and signing a privacy agreement.
Replication code for the analysis is available at \url{https://github.com/adamSales/psGee}.

@@

We estimated principal scores using logistic regression of observed $S$ on school fixed effects and a set of baseline variables chosen with a combination of expert judgement and backwards selection based on the AIC, and evaluated model fit with binned residual plots \citep{gelman2006data}. 
The model was fit to data from students in the ASSISTments group, for whom bottom-out status ($S$) was observed; hence, the same model fit was used for all three treatment contrasts.
The model was fairly successful in distinguishing bottom-outers from non-bottom-outers---the AUC, evaluated using out-of-sample predictions in a 10-fold cross-validation, was roughly 0.795. That is, in roughly 80\% of bottom-outer/non-bottom-outer pairs of subjects, the bottom-outer will have a higher predicted probability from the model.
Coefficient estimates are displayed in the online appendix.

@@

Figure \ref{fig:compare} shows principal effect estimates from \textsc{geepers} compared to analogous estimates using Bayesian mixture modeling and \textsc{psw}, alongside estimates of the overall average treatment effects, labeled ``ATE.''.
%All three sets of estimates relied on the same principal score model described above, in Section \ref{sec:psMod}, though in the mixture model it was fit simultaneously with effect estimation using a Markov Chain Monte Carlo algorithm \citep{rstan}.
Outcome regressions in \textsc{geepers} and the mixture model included all of the covariates that were in the principal score model, in addition to students' pretest scores; also, the models substituted class fixed effects for school fixed effects. Coefficient estimates are displayed in the online appendix.

The Bayesian mixture model assumed that outcomes were normally distributed, conditional on covariates and principal stratum, with a residual standard deviation that varied between principal strata. This modeling assumption is necessarily false, since post-test scores were integers between 0 and 10. All model parameters with support in $\mathbb{R}$ were given standard normal priors, while standard deviation parameters were given half-standard normal priors.
The principal score and outcome models were fit simultaneously using a Markov Chain Monte Carlo algorithm \citep{rstan}.

@@

\section{Discussion}\label{sec:discussion}
\textsc{geepers} is a straightforward approach to principal effect estimation under strong monotonicity, built on widely-used regression models, which is more robust---though sometimes less precise---than alternative approaches under a wide array of scenarios.

There is good reason to hope that extensions to \textsc{geepers}, including cases in which $S$ takes more than two values, may be straightforward.
For instance, in truncation by death problems \citep[e.g.][]{zhangRubin,ding2011} with weak monotonicity, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, interest is typically in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
This scenario is broadly similar to the strong monotonicity, one-way noncompliance situation that this paper discussed.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

\bibliographystyle{plainnat}
\bibliography{MOM}

\clearpage
\setcounter{page}{1}

\begin{center}
\large
    \textbf{Online Appendices for ``GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations''}
\end{center}

\appendix

@@

Replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations $\sum_{i=1}^{n_C}\Psi_i=\bm{0}$ with $\Psi_i=$ 
\begin{equation}\label{eq:estEq1}
\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}
\end{equation}
These are equivalent to the estimating equations for OLS model \eqref{eq:regression0} with $\beta_0=\muc0$, $\beta_1=\muc1-\muc0$, $\beta_2=\mut0-\muc0$, and $\beta_3=\mut1-\mut0-\muc1-\muc0$.
Therefore, under standard OLS regularity conditions the estimated parameter vector $\bm{\hat{\beta}}$ is consistent, completing the proof.

@@

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEstApp}
  \begin{split}
    \heff0_{int}&\equiv \hat{\beta}_2+\bm{\hat{\gamma}_3}'\overline{\bxy}_{Z=1,S=0}\\
    \heff1_{int}&\equiv \hat{\beta}_2+\hat{\beta}_3+(\bm{\hat{\gamma}_3}+\bm{\hat{\gamma}_4})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\heff0_{G}$ and $\heff1_{int}$ are M-estimators, with $\heff0_{int}\rightarrow_p\eff0$ and $\heff1_{G}\rightarrow_p\eff1$ as $n\rightarrow\infty$.
  Under suitible regularity conditions, they are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

@@

\begin{proof}
First of all, by \eqref{eq:interaction}, 
\begin{equation*}
\begin{split}
  \EE[Y_T-Y_C|\st=0]&\\
  =&\EE[Y|Z=1,\st=0]-\EE[Y|Z=1,\st=0]\\
  =&\beta_2+\bm{\gamma_3}'\EE[\bxy|\st=0]
\end{split}
\end{equation*}
and
\begin{equation*}
\begin{split}
  \EE[Y_T-Y_C|\st=1]&\\
  =&\EE[Y|Z=1,\st=1]-\EE[Y|Z=1,\st=1]\\
  =&\beta_2+\beta_3+(\bm{\gamma_3}'+\bm{\gamma_4}')\EE[\bxy|\st=1]
\end{split}
\end{equation*}
Furthermore, $\overline{\bxy}_{Z=1,S=0}\rightarrow \EE[\bxy|\st=0]$ and $\overline{\bxy}_{Z=1,S=1}\rightarrow \EE[\bxy|\st=1]$ as $n\rightarrow \infty$.

@@

First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ (due to randomization)}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}
implying that, according to \eqref{eq:interaction},
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\beta_1\pp+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}\pp\\
                  &=\beta_0+\beta_1R+\bm{\gamma_1}'\bm{X^Y}+\bm{\gamma_2}'\bm{X^Y}R\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_2+(\beta_1+\beta_3)S+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy S\\
    &=\beta_0+\beta_2+(\beta_1+\beta_3)R+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy R
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]=&\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  =&\beta_0+\beta_1\EE[R]+\beta_2\EE[Z]+\beta_3\EE[ZR]+\bm{\gamma_1}'\EE[\bxy]+\bm{\gamma_2}'\EE[\bxy R]\\
  &+\bm{\gamma_3}'\EE[\bxy Z]+\bm{\gamma_4}\EE[\bxy RZ]
\end{align*}

@@

\section{Simulation Results}
\subsection{Plot of  AUC versus $\alpha$}
